Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-05-31 ThesisJP/Background/background.tex
Finished all comments from Sverre on the english text.

The original graphics pipeline involved no conditional branching (caused by if-else control flow statements), and data were typically used once \cite{Purcell2002}. Designers of GPUs could therefore skip advanced features found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the problem of rendering graphics. When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem. How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

@@

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU. This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging systems.

@@

A special kind of instructions called SIMD was mentioned in Section \ref{sec:gpgpu}. These are instructions that are designed to execute a single instruction on multiple data elements in one clock cycle. In the late 1990s, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers. The recently introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply-add (\nom{FMA}{Fused multiply-add (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. The new extension AVX-512, that extends the registers to 512 bit, is found in the Intel accelerator board Xeon Phi, and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014. The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to a theoretical peak arithmetic performance of
\begin{equation}
32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS},
\end{equation}
where FLOPS is short for; floating point operations per second. This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing capabilities to the APU.

Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture from Nvidia provides up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15\times192=2880$ cores. A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)} or SMX).} capable of processing 192 single-precision floating point values per clock cycle per core. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) with close to 1 TFLOPS in computational capabilities. One Kepler GPU therefore provides more than 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared to the SIMD cores on CPUs, result in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge of computer graphics and the graphics pipeline. When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which is maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. While CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction. Because of this the GPU is sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor. This also explain why GPU threads are said to be light-weighted. Such a conceptual difference means that there is an inherent focus on breaking algorithms down into SIMD instructions when programing for the GPU. This often leads to a highly optimized parallel implementation on the GPU and not on the CPU. Manually typed SIMT, or SIMD, instructions is the only way of programming the GPU, while a single thread with automatic vectorization of for-loops is the standard way of programming the CPU.

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something that is not likely to happen in practical applications, and therefore one will usually end up with a throughput which is a fraction of the peak throughput.

The code in Fig.\,\ref{fig:gpu_grid} is an incomplete and minimal example of a GPU program. A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is part of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Therefore, we will probably see a lot of work aimed at reducing this overhead in the future.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in near future, programming the GPU will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction. The relevance of auto-applied GPU instructions will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, GPU programming will hopefully be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation of the human body. Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits. It is also far less costly than other medical imaging modalities and offers real-time imaging. Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle. The tissue separating the left and right ventricle is known as the inter-ventricular septum. The left side is responsible of pumping blood out to the whole body through the aortic valve. It is therefore the strongest and hardest working part of the heart. When blood flows out of the aorta, it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak. The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.

@@

To generate the ultrasonic frequencies used for imaging, a piezoelectric ceramic is used. A piezoelectric material has a special property where mechanical vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. A set of piezoelectric elements organized in an array is referred to as an ultrasound \textit{probe}, and special probes are typically designed for different medical imaging modalities. An example of a probe for medical ultrasound imaging is shown in Fig.\,\ref{fig:us_grid}. %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array, the transmitted pulse will get shaped into a beam\footnote{A beam refers to the accumulated magnitude across time of a pulse traveling in space.} of ultrasound by diffraction as the width (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution in radians of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} are the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed of sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inversely proportional to the pulse length, and is typically better than the lateral resolution. An overall view of the cardiac ultrasound imaging geometry is depicted in Fig.\,\ref{fig:us_grid}. A line in the depicted ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

@@

\subsection{Sampling and processing complexity}
When generating a B-mode image, we have to ensure that the spatial and temporal sampling are sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system that both emits and receives energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector. If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations long. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range (or time)}where $f_c$ is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}. Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. This is far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express, which supports even higher transfer rates, have also recently been introduced.

@@

Software beamforming opens up for cost-efficient and rapid prototyping of advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}) and introduced in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

\subsection{Adaptive beamforming}\label{sec:adaptbf}
Adaptive beamforming has several meanings in the medical ultrasound imaging literature. It can for instance refer to aberration correction \cite{cole1996method} which is the task of applying time delays to compensate for pulse distortion introduced by fatty layers in the human body. However, in this thesis adaptive beamforming will refer to data-dependent weighting derived from a given optimization criterion. The optimization criterion that we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}. A beamformer that uses weights derived with Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}. However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer (Synonym for Capon)}) beamformer \cite{Synnevag2007}, the minimum variance distortionless response beamformer (\nom{MVDR}{Minimum variance distortionless response beamformer (Synonym for Capon)}) [\textbf{Paper I}], or the constrained adaptive beamformer \cite{Mann2002}.

\begin{figure}[t!]
\subfloat[Uniform weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
}
\subfloat[Capon weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
}
\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b). An interfering source is located at 20\degree. Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source. Image courtesy of Jo Inge Buskenes.}
\label{fig:weights}
\end{figure}

>>>>

The original graphics pipeline involved no conditional branching (caused by if-else control flow statements), and data were typically used once \cite{Purcell2002}. Designers of GPUs could therefore skip advanced features found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the problem of rendering graphics. When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem. How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

@@

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle. In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU. This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging systems.

@@

A special kind of instructions called SIMD was mentioned in Section \ref{sec:gpgpu}. These are instructions that are designed to execute a single instruction on multiple data elements in one clock cycle. In the late 1990s, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers. The recently introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply-add (\nom{FMA}{Fused multiply-add (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. The new extension AVX-512, that extends the registers to 512 bit, is found in the Intel accelerator board Xeon Phi, and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014. The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to a theoretical peak arithmetic performance of
\begin{equation}
32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS},
\end{equation}
where FLOPS stands for floating point operations per second. This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing capabilities to the APU.

Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture from Nvidia provides up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15\times192=2880$ cores. A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)} or SMX).} capable of processing 192 single-precision floating point values per clock cycle per core. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) with close to 1 TFLOPS in computational capabilities. One Kepler GPU therefore provides more than 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared to the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge of computer graphics and the graphics pipeline. When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages were needed.  One of the first GPGPU oriented languages was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)} (Compute unified device architecture), a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which is maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. While CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

As we see, a thread has a different conceptual meaning on a GPU than on a CPU.  While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction. Because of this the GPU is sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a single instruction multiple data (SIMD) processor. This also explain why GPU threads are said to be light-weighted. Such a conceptual difference means that there is an inherent focus on breaking algorithms down into SIMD instructions when programing for the GPU. This often leads to a highly optimized parallel implementation on the GPU and not on the CPU. Manually typed SIMT, or SIMD, instructions is the only way of programming the GPU, while a single thread with automatic vectorization of for-loops is the standard way of programming the CPU.

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increases, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something that is not likely to happen in practical applications, and therefore one will usually end up with a throughput which is a fraction of the peak throughput.

The code in Fig.\,\ref{fig:gpu_grid} is an incomplete and minimal example of a GPU program. A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers. This is part of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task. Therefore, we will probably see a lot of work aimed at reducing this overhead in the future.  A recent development is the introduction of \nom{C++ AMP}{C++ Accelerated Massive Parallelism (programming model)} which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in near future, programming the GPU will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions are handled in CPU-code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction. The relevance of auto-applied GPU instructions will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, GPU programming will hopefully be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation of the human body. Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits. It is also far less costly than other medical imaging modalities and offers real-time imaging. Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging. An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}. Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}. Especially note the four chambers of the heart, the left and right atria, and the left and right ventricles. The tissue separating the left and right ventricle is known as the inter-ventricular septum. The left side is responsible for pumping blood out to the whole body through the aortic valve. It is therefore the strongest and hardest working part of the heart. When blood flows out of the aorta, it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak. The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations. When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.

@@

To generate the ultrasonic frequencies used for imaging, a piezoelectric ceramic is used. A piezoelectric material has a special property where mechanical vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. A set of piezoelectric elements organized in an array is referred to as an ultrasound \textit{probe}, and special probes are typically designed for different medical imaging modalities. Examples of probes for imaging of the carotid artery in the neck and for cardiac imaging are shown in Fig.\,\ref{fig:us_grid}, to the left and right respectively. %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element approaches \textit{omnidirectional} characteristics, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array, the transmitted pulse will get shaped into a beam\footnote{A beam refers to the accumulated magnitude across time of a pulse traveling in space.} of ultrasound by diffraction as the width (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution in radians of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} are the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed of sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inversely proportional to the pulse length, and is typically better than the lateral resolution. An overall view of the cardiac ultrasound imaging geometry is depicted in Fig.\,\ref{fig:us_grid}. A line in the depicted ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

@@

\subsection{Sampling and processing complexity}
When generating a B-mode image, we have to ensure that the spatial and temporal sampling is sufficient. We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).  Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system that both emits and receives energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture. The lateral resolution of what is known as an active imaging system is found to be
\begin{align}
\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
\end{align}
\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire 
\begin{align}
N_\theta = \frac{\theta}{\delta\theta}
\end{align}
\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector. If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
\begin{align}
f_{\text{FPS}} = \frac{c}{2rN_\theta}
\end{align} 
images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically have a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations long. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range (or time)}where $f_c$ is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}. Note that after this process the ultrasound data will be complex valued. The total number of samples in a B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. This is far below the peak processing throughput of modern GPUs and CPUs. The main challenge with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express, which support even higher transfer rates, have also recently been introduced.

@@

Software beamforming opens up for cost-efficient and rapid prototyping of advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Papers\,I} and \textbf{II}) and introduced in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Papers\,IV} and \textbf{IX}.

\subsection{Adaptive beamforming}\label{sec:adaptbf}
Adaptive beamforming has several meanings in the medical ultrasound imaging literature. It can for instance refer to aberration correction \cite{cole1996method} which is the task of applying time delays to compensate for pulse distortion introduced by fatty layers in the human body. However, in this thesis adaptive beamforming will refer to data-dependent weighting derived from a given optimization criterion. The optimization criterion that we focus on is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}. A beamformer that uses weights derived with Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}. However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer (Synonym for Capon)}) beamformer \cite{Synnevag2007}, the minimum variance distortionless response beamformer (\nom{MVDR}{Minimum variance distortionless response beamformer (Synonym for Capon)}) [\textbf{Paper I}], or the constrained adaptive beamformer \cite{Mann2002}.

\begin{figure}[t!]
\subfloat[Uniform weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
}
\subfloat[Capon weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
}
\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b). An interfering echo is coming from 20\degree. Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source. Image courtesy of Jo Inge Buskenes.}
\label{fig:weights}
\end{figure}
